---
title: "Comparing E-G predictions with CRISPR data"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: bookdown::html_document2
---

```{r setupDocument, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#save.image("comp.rda")
#stop()
```

```{r comparisonDescription, include=FALSE}
comparison <- snakemake@wildcards$comparison
expt_file <- basename(snakemake@config$comparisons[[comparison]]$expt)
pred_file <- basename(unlist(snakemake@config$comparisons[[comparison]]$pred))
```

<br>

This is the output for the comparison **`r snakemake@wildcards$comparison`**. Following analyses
evaluate how well the experimental data agrees with the predictions of CRE - gene pairs.

Experimental data: **`r expt_file`**  
Predictions: **`r pred_file`**

<br>

***

```{r attachPackages, message=FALSE, warning=FALSE}
# attach required packages and functions
library(tidyverse)
library(here)
library(DT)
library(plotly)
library(ROCR)
library(caTools)
source(file.path(snakemake@scriptdir, "crisprComparisonFunctions.R"))
source(file.path(snakemake@scriptdir, "crisprComparisonPlotFunctions.R"))
```

```{r processData}
# load merged data and config files
merged <- fread(here(snakemake@input$merged))
pred_config <- fread(here(snakemake@input$pred_config))

# filter for valid connections
merged <- subset(merged, IncludeInModel == TRUE)

# add label for each pair based on whether it's significant and activates or represses it's target
merged <- labelPairs(merged, sig_col = "Significant")

# get inverse predictors
inverse_predictors <- pred_config[lowerIsMoreConfident == TRUE, pred.col] %>% 
  outer(snakemake@params$pred_names, ., paste, sep = ".") %>% 
  as.vector(.) %>% 
  c(., "comp.distance") %>% 
  intersect(., colnames(merged))
```

# Effect size vs predictors {.tabset .tabset-pills}
Each predictor listed in the prediction data is plotted against the effect size of enhancer
perturbations reported in the experimental data (e.g. percent change in expression). These plots
show how well a predictor is associated with effects observed in CRISPRi enhancer screens in an
intuitive way.

```{r effectSizeScatter}
# get all cell types in merged
cell_types <- unique(merged$CellType)
if (length(cell_types) > 1) {
  cell_types <- c("combined", cell_types)
}
names(cell_types) <- cell_types

# get all predictors to plot in merged data set
preds_plot <- pred_config$pred.col %>% 
  outer(snakemake@params$pred_names, ., paste, sep = ".") %>% 
  as.vector(.) %>% 
  c(., "comp.distance") %>% 
  intersect(., colnames(merged)) %>% 
  structure(., names = .)

# make scatter plots
es_scatters <- lapply(cell_types, FUN = predScatterPlots, df = merged, predictors = preds_plot,
                      y_col = "EffectSize", ncol_facet = 3, point_size = 2, text_size = 13)
```

```{r saveEffectSizeScatter, include=FALSE}
# calculate plot height based on number of predictors
plot_height <- ceiling(length(preds_plot) / 3) * 3.5

# save scatter plots to files
plotdir <- here(dirname(snakemake@output[[1]]), "plots")
dir.create(plotdir, recursive = TRUE, showWarnings = FALSE)
for (n in names(es_scatters)) {
  ggsave(es_scatters[[n]], filename = here(plotdir, paste0(n, "_EffectSize_scatter_plots.pdf")),
         height = plot_height, width = 10, device = "pdf")
}
```

```{r results='asis', fig.cap="Predictors versus effect size", fig.height=plot_height, fig.width=10}
# print plots for every cell type in tabs
for (i in cell_types){
  cat("##", i, '{-}', '\n', '<br>', '\n')
  plot(es_scatters[[i]])
  cat('\n', '<br>', '\n\n')
}
cat("# {-}")
```

***

<br>

# Precision-Recall {.tabset .tabset-pills}
Precision-recall (PR) curves are used for comparing the performance of different predictors on the
experimental data. The area under the PR curve (AUPRC) provides a single metric of a predictors
performance.

```{r computePRC}
# invert inverse predictors in merged dataset
if (length(inverse_predictors) > 0) {
  merged[, inverse_predictors] <- -1 * merged[, ..inverse_predictors]
}

# compute precision-recall tables for all cell types
pos_col <- snakemake@params$expt_positive_column
pr <- lapply(cell_types, FUN = calcPRCurves, df = merged, predictors = preds_plot,
             pos_col = pos_col) 

# create and export PR summary table (AUC, cutoff, etc)
pr_summary <- lapply(pr, FUN = makePRSummaryTable,
                     min_sensitivity = snakemake@params$min_sensitivity,
                     thresholds = unlist(snakemake@params$pred_threshold))

# convert PR objects to one table per cell type
pr_df <- lapply(pr, FUN = pr2df, calc_f1 = TRUE)

# create output directory for PR data
pr_dir <- here(here(dirname(snakemake@input$merged), "PRC"))
dir.create(pr_dir, recursive = TRUE, showWarnings = FALSE)

# write PR summary tables to files
pr_summary_files <- here(pr_dir, paste0(names(pr_summary), "_pr_summary.txt"))
for (i in seq_along(pr_summary)) {
  write.table(pr_summary[[i]], file = pr_summary_files[[i]], sep="\t", quote = FALSE,
              row.names = FALSE)
}

# write PR tables to files
pr_files <- here(pr_dir, paste0(names(pr_df), "_pr_curve.txt"))
for (i in seq_along(pr_df)) {
  write.table(pr_df[[i]], file = pr_files[[i]], sep="\t", quote = FALSE, row.names = FALSE)
}
```

```{r plotPRC, fig.height=4.5, fig.width=7, warning=FALSE}
# assign prediction class to merged dataset
merged <- addPredictionClassLabels(merged, perf_summary = pr_summary[[1]], pos_col = pos_col)

# write annotated dataset to output file
merged_annotated_file <- here(dirname(snakemake@input$merged), "expt_pred_merged_annotated.txt")
write.table(merged, merged_annotated_file, sep = "\t", quote = FALSE, row.names = FALSE)

# calculate percentage of true positives in dataset per cell type
pct_pos <- vapply(cell_types, FUN = calcPctPos, df = merged, pos_col = pos_col,
                  FUN.VALUE = numeric(1))

# make PRC plots
pr_plots <- mapply(
  FUN = makePRCurvePlot, pr_df, pct_pos,
  MoreArgs = list(pred_cols = preds_plot, plot_name = "PRC full experimental data",
                  point_size = 3, text_size = 15, 
                  min_sensitivity = snakemake@params$min_sensitivity),
  SIMPLIFY = FALSE
  )

# set custom color for PR curves
pr_plots <- lapply(pr_plots, FUN = function(p) {
  p + scale_colour_brewer(type = "qual", palette = "Set1")
})

# save plots to file
pr_plot_files <-  here(plotdir, paste0(names(pr_plots), "_prc.pdf"))
for (i in seq_along(pr_plots)) {
  ggsave(pr_plots[[i]], filename = pr_plot_files[[i]], height = 4.5, width = 7, device = "pdf")
}
```

Precision-recall performance of each predictor:

```{r results='asis', fig.cap="Precision-recall curves", warning=FALSE}
# print plots for every cell type in tabs
for (i in cell_types){
  cat("##", i, '{-}', '\n', '<br>', '\n')
  plot(pr_plots[[i]])
  cat('\n', '<br>', '\n\n')
}
cat("# {-}")
```

This table shows a precision-recall summary for each predictor:
```{r, fig.cap = "Precision-recall summary"}
# combine PR summary into one table for display
pr_summary_df <- bind_rows(pr_summary, .id = "cell_type")

# pretty print PR summary table
datatable(pr_summary_df, options = list(pageLength = 20), autoHideNavigation = TRUE)
```

***
